---
title: 'Assignment 9: Spatial Analysis in R'
author: "Kamil Burak Karayel"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on spatial analysis.

## Directions
1. Rename this file `<FirstLast>_A09_SpatialAnalysis.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
4. Work through the steps, **creating code and output** that fulfill each instruction.
5. Be sure to **answer the questions** in this assignment document. Space for your answers is provided in this document and is indicated by the ">" character. If you need a second paragraph be sure to start the first line with ">". You should notice that the answer is highlighted in green by RStudio.
5. When you have completed the assignment, **Knit** the text and code into a single **HTML** file.


## DATA WRANGLING

### Set up your session

1.    Import libraries: tidyverse, sf, leaflet, here, and mapview
2.    Execute the `here()` command to display the current project directory

```{r Prepare workspace, error=F, warning=F}
#1. 
library(tidyverse)
library(sf)
library(leaflet)
library(here)
#install.packages("mapview")
library(mapview); mapviewOptions(fgb = FALSE)

#2. 
getwd()
here()

```


### Read (and filter) county features into an sf dataframe and plot

In this exercise, we will be exploring stream gage height data in Nebraska corresponding to floods occurring there in 2019. First, we will import from the US Counties shapefile we've used in lab lessons, filtering it this time for just Nebraska counties. Nebraska's state FIPS code is `31` (as North Carolina's was `37`).

3.    Read the `cb_2018_us_county_20m.shp` shapefile into an sf dataframe, filtering records for Nebraska counties (State FIPS = 31)
4.    Reveal the dataset's coordinate reference system
5.    Plot the records as a map (using `mapview` or `ggplot`)

```{r Read the county data into an sf dataframe}
#3. Read in Counties shapefile into an sf dataframe, filtering for just NE counties
#Filter Nebraska where its FIPS code is 31.
Q3 <- st_read('./Data/Spatial/cb_2018_us_county_20m.shp') %>% filter(STATEFP==31)


#4. Reveal the CRS of the counties features
st_crs(Q3)


#5. Plot the data
#Counties in Nebraska assigned different colors based on county area of land.
mapView(Q3, zcol= "ALAND", layer.name="Area")

```

6.    What is the EPSG code of the Counties dataset? Is this a geographic or a projected coordinate reference system? (In other words, does this CRS use angular or planar coordinate units?) To what datum is this CRS associated? (Tip: lookup the EPSG code on <https://epsg.io> or <https://spatialreference.org>)

> ANSWER: 
> By revealing the CRS of the counties dataset we can see that EPSG code is 4269 (ID["EPSG",4269]). EPSG code 4269 uses The North American Datum of 1983 (NAD 83) as its geographic coordinate reference system (GCRS). It is the horizontal and geometric control datum for the United States, Canada, Mexico, and Central America. The dataset uses angular coordinate units in degrees of latitude and logitude.


### Read in gage locations csv as a dataframe, then display the column names it contains

Next we'll read in some USGS/NWIS gage location data added to the `Data/Raw` folder. These are in the `NWIS_SiteInfo_NE_RAW.csv` file.(See `NWIS_SiteInfo_NE_RAW.README.txt` for more info on this dataset.)

7.    Read the `NWIS_SiteInfo_NE_RAW.csv` file into a standard dataframe, being sure to set the `site_no` field as well as other character columns as a factor. 

8.    Display the structure of this dataset.

```{r Read in USGS Gage Locations}
#7. Read in gage locations csv as a dataframe
#site_no consists of numbers, and stringsAsFactors doesn't make it factor. So I manually changed it.
Q7 <- read.csv("./Data/Raw/NWIS_SiteInfo_NE_RAW.csv", 
                             stringsAsFactors = TRUE, 
                             colClasses = c("site_no" = "factor"))

#8. Display the structure of the dataframe
str(Q7)
#now every column except numbers are factors.

```

9.    What columns in the dataset contain the x and y coordinate values, respectively?  
> ANSWER:
> "dec_long_va" contains x coordinate values and "dec_lat_va" contains y coordinate values.


### Convert the dataframe to a spatial features ("sf") dataframe

10.   Convert the dataframe to an sf dataframe.
    *   Note: These data use the same coordinate reference system as the counties dataset

11. Display the structure of the resulting sf dataframe

```{r Convert dataframe to sf dataframe}
#10. Convert to an sf object
#data from Q7, x is dec_long_va, y is dec_lat_va, CRS is 4269 as the counties dataset.
Q10 <- st_as_sf(Q7, coords = c("dec_long_va", "dec_lat_va"), crs = 4269)


#11. Display the structure
str(Q10)


```

12.   What new field(s) appear in the sf dataframe created? What field(s), if any, disappeared?

>ANSWER: 
>Deleted "dec_long_va" and "dec_lat_va". A new field "geometry" added as x and y points from deleted columns.


### Plot the gage locations on top of the counties

13. Use `ggplot` to plot the county and gage location datasets.
  *   Be sure the datasets are displayed in different colors
  *   Title your plot "NWIS Gage Locations in Nebraska"
  *   Subtitle your plot with your name

```{r Plot the spatial features}
#13. Plot the gage locations atop the county features
Q13 <- ggplot() +
  geom_sf(data = Q3, color='blue') +    #county dataset will be blue. 
  geom_sf(data = Q10, color='red') +    #gage locations dataset will be red
  labs(title = "NWIS Gage Locations in Nebraska", 
       subtitle = "Kamil Burak Karayel")

Q13

```


### Read in the gage height data and join the site location data to it.

Lastly, we want to attach some gage height data to our site locations. I've constructed a csv file listing many of the Nebraska gage sites, by station name and site number along with stream gage heights (in meters) recorded during the recent flood event. This file is titled `NWIS_SiteFlowData_NE_RAW.csv` and is found in the Data/Raw folder.

14. Read the `NWIS_SiteFlowData_NE_RAW.csv` dataset in as a dataframe
    *   Pay attention to which fields should be imported as factors!
15. Show the structure of the dataframe.
16. Join our site information (already imported above) to these gage height data
    *   The `site_no` and `station_nm` can both/either serve as joining attributes
    *   Construct this join so that the result only includes records features where both tables have data (N=136)
17. Show the column names of this resulting spatial dataframe
18. Show the dimensions of the resulting joined dataframe

```{r Read in gage site data}
#14. Read the site flow data into a data frame
Q14 <- read.csv("./Data/Raw/NWIS_SiteFlowData_NE_RAW.csv", 
                stringsAsFactors = TRUE,    #change characters to factor
                colClasses = c("site_no" = "factor"))  #change class of "site_no" to factor


#15. Show the structure of the dataframe 
str(Q14)   
#now classes of site_no, station_nm, date are factor, gage_ht is number.


#16. Join the flow data to our NWIS gage location spatial dataframe
#merge two datasets with respect to site_no
Q16 <- merge(x= Q14,
             y= Q10,
             by.x = "site_no",
             by.y="site_no")


#17. Show the column names in the resulting spatial dataframe
colnames(Q16)


#18. Show the dimensions of this joined dataset
dim(Q16)
#it is 136x9

```

### Map the pattern of gage height data

Now we can examine where the flooding appears most acute by visualizing gage heights spatially.
19.   Plot the gage sites on top of counties (using `mapview`, `ggplot`, or `leaflet`) 
    *   Show the magnitude of gage height by color, shape, other visualization technique.

```{r map gage sites}
#Map the points, sized by gage height
Q19 <- ggplot() +
  geom_sf(data = Q3, aes(), color = 'blue') +   #counties will be blue
  geom_sf(data = Q16, aes(geometry = geometry, color = gage_ht, shape = "*")) +
  #gage locations will be colored by height.
  labs(title = "Gage sites in Nebraska", 
       subtitle = "Color scale indicates the water level")

Q19

```

---

## SPATIAL ANALYSIS

Up next we will do some spatial analysis with our data. To prepare for this, we should transform our data into a projected coordinate system. We'll choose UTM Zone 14N (EPGS = 32614).


### Transform the counties and gage site datasets to UTM Zone 14N

20. Transform the counties and gage sf datasets to UTM Zone 14N (EPGS = 32614).
21. Using `mapview` or `ggplot`, plot the data so that each can be seen as different colors

```{r spatial analysis}
#20 Transform the counties and gage location datasets to UTM Zone 14
#counties dataset transformed from NAD83 to UTM Zone 14 as EPSG will be 32614.
Q20_counties <- st_transform(Q3,32614)
#gage locations dataset transformed from NAD83 to UTM Zone 14 as EPSG will be 32614.
Q20_gages <- st_transform(Q10,32614)

#21 Plot the data
mapview(Q20_counties, col.region = "orange", layer.name= "Counties") +
mapview(Q20_gages, col.region = "lightblue", layer.name= "Gage sites")


```


### Select the gages falling within a given county

Now let's zoom into a particular county and examine the gages located there. 
22. Select Saunders county from your projected county sf dataframe 
23. Select the gage sites falling within that county to a new spatial dataframe
24. Select the gage sites within 15km of the county to a new spatial dataframe
25. Create a plot showing (each symbolized distinctly):
 * all Nebraska counties,
 * the selected county, 
 * the gage sites in that county,
 * and the gage sites within 15 km of the county


```{r spatial selection}
#22 Select the county
#include only Saunders County.
Q22 <- Q20_counties %>% 
  filter(NAME == "Saunders")

#23 Spatially select gages within the selected county
#show gages in only Saunders.
Q23 <- st_intersection(Q20_gages, Q22)

mapview(Q22, col.region = "orange", layer.name= "Saunders County")+
mapview(Q23, col.region = "lightgreen", layer.name= "Gage sites")

#24 Spatially select gages within 15 k of the selected county
#create a buffer zone outside of Saunders with 15000 m of diameter.
Q24_buffer <- st_buffer(Q22, 15000)
#show gages in Saunders County and buffer zone.
Q24 <- st_intersection(Q20_gages, Q24_buffer)


#25 Plot
mapview(Q24_buffer, col.region = "orange", layer.name= "Saunders County with 15km Buffer Zone")+
  mapview(Q24, col.region = "purple", layer.name= "Gage sites")


```
